Integration of national demographic-disturbance relationships into local caribou population viability projections can reduce uncertainty and inform monitoring decisions

### 2024-08-27 analysis/paper/paper.Rmd

As suggested by Dunham et al. [-@dunham_strategic_2023] we also
calculate the expected value of perfect information (EVPI) and the
expected value of sample information (EVSI) [@runge_which_2011;
@yokota_value_2004]. We assign a value or utility of
$\nu_{\alpha,\psi}=1$ when the true state of the population $\psi$
matches the assigned status $\alpha$, and value of $\nu_{\alpha,\psi}=0$
when the status assessment is incorrect. This is "the value ascribed by
the decision-maker to the outcome, and is thus the measure of management
performance" [@runge_which_2011]. The expected value of current
information is the average utility of decisions informed by knowledge of
the disturbance scenario $\tau$ (Fig \@ref(fig:figDisturbance)) and
priors from the national model:
\begin{equation}
EV_{current} = max_{\alpha}\sum_\psi[P(\psi|\tau)\nu_{\alpha,\psi}],
  (\#eq:EVcurrent)
\end{equation}
where $P(\psi|\tau)$ is the prior probability that the
true state of the population is $\psi$, and each population is assigned
a status $\alpha$ informed by the prior. The expected value under
certainty (i.e. the expected value of a decision when the true state of
the population is known) is $EV_{certainty}=1$. The difference between
$EV_{certainty}$ and $EV_{current}$ gives EVPI, a measure of the maximum
possible benefit that could theoretically be gained if the system were
perfectly known.

@@

Analysis of simulated example demographic trajectories confirms that our
Bayesian model can effectively combine local demographic data with
priors from national demographic-disturbance relationships to reduce
uncertainty in boreal caribou demographic projections (Figs
\@ref(fig:examplePredictions) and \@ref(fig:figDiffs)). As expected,
uncertainty about the population growth rate decreases as sampling
effort in each year (i.e. number of collars) or the number of years of
monitoring (Fig \@ref(fig:figDiffs)) increase. Uncertainty is higher
when the forecast horizon is long (compare 5 and 20 year projections in
Fig \@ref(fig:figDiffs)). Uncertainty is also higher in our high
anthropogenic disturbance scenarios (Fig \@ref(fig:figDiffs)) because
stochastic variation in population growth rates increase as populations
become very small (as in top right panel of Fig
\@ref(fig:examplePredictions)).

Our analysis becomes more informative when we examine the probability of
correctly assessing population status (Fig \@ref(fig:figStatusErrors))
and the expected value of information (Figs \@ref(fig:figEVSI) and E.1).
When anthropogenic disturbance is high (right panels of Figs
\@ref(fig:figStatusErrors), \@ref(fig:figEVSI), and E.1)), projections
of population status informed only by priors from the national
demographic-disturbance relationships are accurate, and adding
additional local monitoring data does not improve accuracy.
Interestingly, when the projection period is long and anthropogenic
disturbance is high, we see a negative EVSI (lower right panel of Fig
\@ref(fig:figEVSI)), and we do not see an associated decrease in the
probability of correctly assessing population status (lower right panel
of Fig \@ref(fig:figStatusErrors)). This means that the long-term
projections informed by local monitoring data are on average less
precise but not less accurate, which can happen when there is very small
prior probability that population is viable (orange band in the top
right panel of Fig \@ref(fig:examplePredictions) falls below 1). At
lower levels of disturbance, local monitoring can reduce uncertainty,
but there are diminishing returns as the number of collars deployed or
the number of years of monitoring increases (Figs
\@ref(fig:figStatusErrors) and \@ref(fig:figEVSI)). The shapes of the
relationships between monitoring effort and the probability of correct
status assessment depend on how anthropogenic disturbance is changing
over time, but there are diminishing benefits of increased monitoring
regardless of whether disturbance is increasing or decreasing
(Supplement D). Adding unknown bias increases uncertainty, but these
increases are not large for the set of parameters we investigate
(Supplement F).

>>>>

As suggested by Dunham et al. [-@dunham_strategic_2023] we also
calculate the expected value of perfect information (EVPI) and the
expected value of sample information (EVSI) [@runge_which_2011;
@yokota_value_2004]. We assign a value or utility of
$\nu_{\alpha,\psi}=1$ when the true state of the population $\psi$
matches the assigned status $\alpha$, and value of $\nu_{\alpha,\psi}=0$
when the status assessment is incorrect. This is "the value ascribed by
the decision-maker to the outcome, and is thus the measure of management
performance" [@runge_which_2011]. The expected value of current
information is the average utility of decisions informed by knowledge of
the disturbance scenario $\tau$ (Figure \@ref(fig:figDisturbance)) and
priors from the national model:
\begin{equation}
EV_{current} = max_{\alpha}\sum_\psi[P(\psi|\tau)\nu_{\alpha,\psi}],
  (\#eq:EVcurrent)
\end{equation}
where $P(\psi|\tau)$ is the prior probability that the
true state of the population is $\psi$, and each population is assigned
a status $\alpha$ informed by the prior. The expected value under
certainty (i.e. the expected value of a decision when the true state of
the population is known) is $EV_{certainty}=1$. The difference between
$EV_{certainty}$ and $EV_{current}$ gives EVPI, a measure of the maximum
possible benefit that could theoretically be gained if the system were
perfectly known.

@@

Analysis of simulated example demographic trajectories confirms that our
Bayesian model can effectively combine local demographic data with
priors from national demographic-disturbance relationships to reduce
uncertainty in boreal caribou demographic projections (Figures
\@ref(fig:examplePredictions) and \@ref(fig:figDiffs)). As expected,
uncertainty about the population growth rate decreases as sampling
effort in each year (i.e. number of collars) or the number of years of
monitoring (Figure \@ref(fig:figDiffs)) increase. Uncertainty is higher
when the forecast horizon is long (compare 5 and 20 year projections in
Figure \@ref(fig:figDiffs)). Uncertainty is also higher in our high
anthropogenic disturbance scenarios (Figure \@ref(fig:figDiffs)) because
stochastic variation in population growth rates increase as populations
become very small (as in top right panel of Figure
\@ref(fig:examplePredictions)).

Our analysis becomes more informative when we examine the probability of
correctly assessing population status (Figure \@ref(fig:figStatusErrors))
and the expected value of information (Figures \@ref(fig:figEVSI) and E.1).
When anthropogenic disturbance is high (right panels of Figures
\@ref(fig:figStatusErrors), \@ref(fig:figEVSI), and E.1)), projections
of population status informed only by priors from the national
demographic-disturbance relationships are accurate, and adding
additional local monitoring data does not improve accuracy.
Interestingly, when the projection period is long and anthropogenic
disturbance is high, we see a negative EVSI (lower right panel of Figure
\@ref(fig:figEVSI)), and we do not see an associated decrease in the
probability of correctly assessing population status (lower right panel
of Figure \@ref(fig:figStatusErrors)). This means that the long-term
projections informed by local monitoring data are on average less
precise but not less accurate, which can happen when there is very small
prior probability that population is viable (orange band in the top
right panel of Figure \@ref(fig:examplePredictions) falls below 1). At
lower levels of disturbance, local monitoring can reduce uncertainty,
but there are diminishing returns as the number of collars deployed or
the number of years of monitoring increases (Figures
\@ref(fig:figStatusErrors) and \@ref(fig:figEVSI)). The shapes of the
relationships between monitoring effort and the probability of correct
status assessment depend on how anthropogenic disturbance is changing
over time, but there are diminishing benefits of increased monitoring
regardless of whether disturbance is increasing or decreasing
(Supplement D). Adding unknown bias increases uncertainty, but these
increases are not large for the set of parameters we investigate
(Supplement F).
